---
title: "Mapping Data with TidyCensus"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(tidycensus)
library(ggiraph)
library(mapview)
library(leaflet)
library(mapgl)
options(scipen = 999)
```

### Welcome!

This notebook is all about using the most popular/modern packages to map data in R. They all follow the same style but have their unique functions and strengthens/weaknesses. With these options you should able to design almost any map you can think of. There are many more packages not included here but this is a great start. Here's the documentation for each:

- [mapview](https://r-spatial.github.io/mapview/)

- [ggplot](https://ggplot2.tidyverse.org/reference/geom_map.html)

- [ggiraph](https://davidgohel.github.io/ggiraph/reference/geom_map_interactive.html)

- [leaflet](https://rstudio.github.io/leaflet/index.html)

- [mapgl](https://walker-data.com/mapgl/)

If our collective dream is to make maps, we're gonna need some data. There are several file types and data formats that need a few key tricks to "clean" in order to map in R but we'll get to that later. For now let's use the awesome [tidycensus](https://walker-data.com/tidycensus/) package to bring in some census data straight to R in a format that is map-ready.

### API Key

First, you'll need to set up an api key though:\
A key can be obtained for free from <https://api.census.gov/data/key_signup.html>.

Then paste the api key into you `.Renviron` by typing in the console:

`usethis::edit_r_environ()`

This will open a file for you to paste your key into exactly like this:

CENSUS_API_KEY='Your Key Here'

Save the file and **RESTART R** for the key to work

### Load in Data

Now, we can load our data in! Let's start with a median income (*code B19013_001*) for Honolulu county

```{r}
hnl <- get_acs(
  state = "HI",
  county = "Honolulu",
  geography = "tract",
  variables = "B19013_001",
  geometry = TRUE,
  year = 2024
) |>
  dplyr::filter(GEOID != "15003981200") # northwestern HI island filtered out

```

Great, 329 census tracts with estimates of median income. Before mapping let's use ggplot to look at median income distribution.

### Plot Data

```{r}
ggplot(data = hnl, aes(x = estimate)) +
  geom_histogram()
```

We can dress this up a bit

```{r}
ggplot(data = hnl, aes(x = estimate)) +
  geom_histogram(color = "black", bins = 50) +
  theme_minimal() + 
  geom_vline(xintercept = median(hnl$estimate, na.rm = TRUE), col ="purple") +
  labs(x = "Median Income is $USD")
```

```{r}
ggplot(data = hnl, aes(x = estimate)) +
  geom_area(stat = "density", fill = "purple") +
  theme_minimal() +
  labs(x = "Median Income is $USD")
```

Looks like a fairly normal distribution around \$120,000. We can get some more stats

```{r}
summary(hnl$estimate)
```

AI Suggestion: **ECDF (empirical cumulative distribution)** — great for answering "what % of tracts are below \$X?"

```{r}
ggplot(hnl, aes(x = estimate)) +
  stat_ecdf() +
  theme_minimal() +
  labs(y = "Cumulative proportion of tracts",
       x = "Median Income is $USD")
```

Now that we have a good understanding of the distribution of our Median Income variable, let's make some maps. We'll start with the simplest map for each package then improve it.

### Map Data

#### *Mapview*

Mapview is awesome because you can get a nice, interactive map in one line of code (4 words)

```{r}
mapview(hnl, zcol = "estimate")
```

```{r}
mapview(
  hnl, 
  zcol = "estimate", 
  col.regions = viridisLite::magma(100),
  layer.name = "Median Income ($)",
  alpha.regions = 0.8,
  label = hnl$NAME
)
```

You can also map breweries as a bonus

```{r}
mapview(breweries)
```

#### ggplot

ggplot is is the most common plotting package in R for a reason. We used it above for other visualization, let's use it for mapping

```{r}
ggplot(data = hnl, aes(fill = estimate)) + 
  geom_sf()
```

```{r}
ggplot(data = hnl, aes(fill = estimate)) + 
  geom_sf(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    option = "viridis", 
    labels = scales::label_dollar(),
    na.value = "grey90"
  ) +
  theme_minimal() +
  labs(
    title = "Median Household Income",
    subtitle = "By Census Tract",
    fill = "Median Income"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

Very Nice, we can also make these interactive with ggiraph

```{r}
medincome_map <- ggplot(data = hnl, aes(fill = estimate)) +
  geom_sf_interactive(
    aes(tooltip = paste0("Estimate: $", estimate, "\nTract: ", NAME))
  ) +
  scale_fill_viridis_c(option = "magma") +
  theme_minimal()

girafe(
  ggobj = medincome_map,
  options = list(
    opts_zoom(min = 1, max = 5)
  )
)
```

#### Leaflet

Leaflet has been a long-time staple in the R mapping community but takes a bit more code to look nice. No simple `leaflet(hnl)`

```{r}
pal <- colorNumeric("magma", domain = hnl$estimate, na.color = "transparent")

leaflet(hnl) |>
  addPolygons(
    fillColor = ~pal(estimate),
    fillOpacity = 0.8,
    color = "white",
    weight = 0.5,
    label = ~paste0(NAME, ": $", scales::comma(estimate))
  ) |>
  addLegend(pal = pal, values = ~estimate, title = "Median Income", opacity = 0.8)
```

#### Mapgl

mapgl is the most recent package we are using built by the very talented GIS master Kyle Walker. It also requires a bit more code but is highly modern, functional, and customizable

```{r}
# Automatic continuous scale with equal breaks
continuous_scale <- interpolate_palette(
  data = hnl,
  column = "estimate",
  method = "equal",
  n = 5,
  palette = viridisLite::cividis
)

mapboxgl(bounds = hnl) |>
  add_fill_layer(
    id = "income_tracts",
    source = hnl,
    fill_color = continuous_scale$expression,
    fill_opacity = 0.8
  ) |>
  add_legend(
    "Median Household Income",
    values = get_legend_labels(continuous_scale, digits = 0),
    colors = get_legend_colors(continuous_scale),
    type = "continuous",
    position = "bottom-right",
    width = "220px"
  )
```

### Challenges

Please complete the following challenges with code below the instructions, save your file, commit, and "push" to github. We can cover how to do this in class.

Acceptable commit messages:

- "Mapping in R is all-time"

- " I want to map in R forever"

- "Amber and Connor have taught me everything I know"

Unacceptable commit messages:

- "AI did this for me"

- "commit"

#### Challenge 1

- Pick a county in the U.S. (not honolulu)

- Read in data for median age with `get_acs()`

  - `median age = "B01002_001"`

- Make a histogram or area plot with `ggplot()`

- Note the min, max, mean, median and quartiles with `summary()`

- Make a map of median age by census tract for your chosen county with ggplot

  Answers:

```{r}
#Read in data for median age with get_acs()

UdaQueen <- get_acs(
  state = "NY",
  county = "Queens",
  geography = "tract",
  variables = "B01002_001",
  geometry = TRUE,
  year = 2023
)

```

```{r}
#Make a histogram or area plot with ggplot()

ggplot(data = UdaQueen, aes(x = estimate)) +
  geom_histogram(color = "black", bins = 30) +
  theme_minimal() + 
  geom_vline(xintercept = median(UdaQueen$estimate, na.rm = TRUE), col ="purple") +
  labs(
    title = "Median age by census tract for Queens county, NY",
    x = "NY Median age",
    y = "Number of Census tract"
  )

#Note the min, max, mean, median and quartiles
summary(UdaQueen$estimate)

#Make a map of median age by census tract for your chosen county with ggplot
ggplot(data = UdaQueen, aes(fill = estimate)) + 
  geom_sf(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    option = "magma", 
    labels = scales::label_dollar(),
    na.value = "grey90"
  ) +
  theme_minimal() +
  labs(
    title = "Median age for queen county, NY",
    subtitle = "By Census Tract",
    fill = "Median Age"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

- Read in data for median income for the same county you chose

  ```         
   median_income = "B19013_001"
  ```

- Make a histogram or area plot with `ggplot()`

- Note the min, max, mean, median and quartiles with `summary()`

- Make a map of median income by census tract for your chosen county with `mapview()`

Answers:

```{r}
IncomeQueen <- get_acs(
  state = "NY",
  county = "Queens",
  geography = "tract",
  variables = "B19013_001",
  geometry = TRUE,
  year = 2023
)
```

```{r}
ggplot(data = IncomeQueen, aes(x = estimate)) +
  geom_histogram(color = "black", bins = 20) +
  theme_minimal() + 
  geom_vline(xintercept = median(IncomeQueen$estimate, na.rm = TRUE), col ="purple") +
  labs(
    title = "Median Income by census tract for Queens county, NY",
    x = "NY Median Income",
    y = "Number of Census tract"
  )

#Note the min, max, mean, median and quartiles
summary(IncomeQueen$estimate)

#Make a map of median income by census tract for your chosen county with mapview()

mapview(
  IncomeQueen, 
  zcol = "estimate", 
  col.regions = viridisLite::magma(100),
  layer.name = "Median Income ($)",
  alpha.regions = 0.8,
  label = IncomeQueen$NAME
)
```

Make a scatterplot with `ggplot` showing the relationship between median age and median income between census tracts for your chosen county

#### Challenge 2

[Follow this guide](https://walker-data.com/mapgl/articles/map-design.html#bivariate-styling:~:text=80-,Bivariate%20styling,-Bivariate%20maps%20visualize) to make a bivariate map of median age and income for your chosen county

Answers:

```{r}

Queens_age_income <- get_acs(
  geography = "tract",
  variables = c(
    median_age = "B01002_001",
    median_income = "B19013_001"
  ),
  state = "NY",
  county = "Queens",
  year = 2023,
  geometry = TRUE,
  output = "wide"
)

names(Queens_age_income) <- sub("E$", "", names(Queens_age_income))

Queens_age_income$popup <- glue::glue(
  "<strong>GEOID: </strong>{Queens_age_income$GEOID}<br>",
  "<strong>Median age: </strong>{round(Queens_age_income$median_age, 1)}<br>",
  "<strong>Median household income: </strong>${format(round(Queens_age_income$median_income), big.mark = ',')}"
)

age_income_scale <- bivariate_scale(
  data = Queens_age_income,
  x = "median_income",
  y = "median_age", na_color = "white", palette = "blue_red"
)

maplibre(
  style = carto_style("positron"),
  bounds = Queens_age_income
) |>
  add_fill_layer(
    id = "Queens_age_income",
    source = Queens_age_income,
    fill_color = age_income_scale$expression,
    fill_opacity = 0.75,
    fill_outline_color = "rgba(255,255,255,0.25)",
    popup = "popup",
    hover_options = list(
      fill_opacity = 1
    )
  ) |>
  add_bivariate_legend(
    age_income_scale,
    legend_title = " Queens Age and income",
    x_title = "Higher income",
    y_title = "Older median age",
    layer_id = "Queens_age_income",
    style = legend_style(
      background_color = "white",
      background_opacity = 0.94,
      border_color = "#d1d5db",
      border_width = 1,
      shadow = TRUE
    )
  )
```
