---
title: "🗺️ Interactive Mapping with mapgl and Shiny"
author: "DS421 – Cartographic Design"
format:
  html:
    code-fold: true
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

## 1️⃣ Census Shiny App

#### Load Libraries

```{r}
library(shiny)
library(tidyverse)
library(sf)
library(mapgl)
library(lubridate)
library(bslib)
library(here)
library(tidycensus)
library(ggpubr)
library(plotly)
library(mapview)
library(leaflet)
```

#### View Census Variables

```{r}
acs_vars <- load_variables(year = 2023, dataset = "acs5", cache = TRUE)
```

#### Transform Data

```{r}
# Define raw ACS variable codes
raw_vars <- c(
  median_income = "B19013_001",
  bachelors_count = "B15003_022",  # Count, not percent
  median_gross_rent = "B25064_001",
  total_population = "B01003_001"
)

# Labels for Shiny dropdown (key = column name in wide data)
variable_choices <- c(
  "Median Household Income" = "median_income",
  "Percent with Bachelor's Degree" = "pct_bachelors",
  "Median Gross Rent" = "median_gross_rent"
)

# Pull and transform ACS data
hawaii_data_raw <- get_acs(
  geography = "tract",
  state = "HI",
  variables = raw_vars,
  geometry = TRUE,
  year = 2023
) |>
  filter(GEOID != "15003981200") |>  # Remove Kalawao County
  st_transform(4326)

# Convert to wide format + calculate % with bachelor's
hawaii_data <- hawaii_data_raw |>
  select(GEOID, NAME, variable, estimate, geometry) |>
  st_drop_geometry() |>   # temp drop for pivot
  pivot_wider(names_from = variable, values_from = estimate) |>
  mutate(
    pct_bachelors = 100 * bachelors_count / total_population
  ) |>
  left_join(
    hawaii_data_raw |> select(GEOID, geometry) |> distinct(),
    by = "GEOID"
  ) |>
  st_as_sf()
```

## 🧩 Challenge 1

Can you make a histogram of median rent colored by county. Hint: You will have to make a new column called county based on the values in the NAME column

```{r}
ggplot(data = hawaii_data, aes(x = median_gross_rent)) +
  geom_histogram()
```

```{r}
hawaii_data <- hawaii_data %>%
  mutate(
    county = str_split_fixed(NAME, "; ", 3)[,2]
  )
ggplot(data = hawaii_data,
       aes(x= median_gross_rent, fill = county))+
  geom_histogram()
```

## 🧩 Challenge 2

A. Can you make a scatterplot showing the relationship between median income and median gross rent. Can you use the `stat_cor()`function to show a correlation coefficient and p value

```{r}
scatter_plot <- ggplot(hawaii_data, aes(x = median_income, y = median_gross_rent)) +
  geom_point() +
  stat_cor() +
  geom_smooth()

```

```{r}

```

B. Can you make this scatterplot interactive with `ggplotly()`

```{r}
ggplotly(scatter_plot)
```

## 🧩 Challenge 3

Can you make a choropleth map of Hawaii using `mapview()` that shows median rent by census tract

```{r}
mapview(hawaii_data, zcol = "median_gross_rent")
```

#### Building a Shiny App

We have these 3 variables in hawaii_data, it would be nice if users can select a variable from a dropdown: [Introducing Shiny](https://shiny.posit.co/)

There are many mapping packages we can use with Shiny. The simplest for now is `mapview`. It is good to note that the mapgl package is also designed specifically to [work with advanced shiny apps](https://walker-data.com/mapgl/articles/shiny.html).

```{r}

ui <- fluidPage(
  titlePanel("Hawaii Census Map"),
  sidebarLayout(
    sidebarPanel(
      selectInput("var", "Choose a variable to map:", choices = variable_choices)
    ),
    mainPanel(
      leafletOutput("map", height = "700px")
    )
  )
)

server <- function(input, output, session) {
  output$map <- renderLeaflet({
    mapview(
      hawaii_data,
      zcol = input$var,
      layer.name = names(variable_choices)[variable_choices == input$var]
    )@map
  })
}

shinyApp(ui, server)
```

## 🧩 Challenge 4

Can you create a shiny app using your LANDSAT_USA_contiguous_and_Hawaii_7d.csv in the data folder that allows users to filter the map by confidence. For example, the user can select from a side dropdown to only show *high* confidence fires, *medium* confidence fires, or *low* confidence fires. Do not use AI for this and use the example above. AI Bonus: Add a date slider so users can toggle the map by date. Can you add a heatmap layer users can toggle on and off?

```{r}
wildfire_df <- read.csv("data/LANDSAT_USA_contiguous_and_Hawaii_7d.csv", header = TRUE)
```

```{r}
wildfire_sf <- st_as_sf(
  wildfire_df,
  coords = c("longitude", "latitude"),
  crs = 4326
)
```

```{r}
mapview(wildfire_df, xcol = "longitude", ycol = "latitude", zcol = "confidence", crs = 4326)
```

```{r}
variable_choices <- c(
  "low" = "low",
  "medium" = "medium",
  "high" = "high")
```

```{r}
library(shiny)
library(leaflet)
library(mapview)
library(sf)


wildfire_sf <- st_as_sf(
  wildfire_df,
  coords = c("longitude", "latitude"),
  crs = 4326
)


variable_choices <- c("daynight", "scan", "track") 

ui <- fluidPage(
  titlePanel("Hawaii Wildfire Map"),
  sidebarLayout(
    sidebarPanel(
      selectInput("var", "Choose variable to map", choices = variable_choices)
    ),
    mainPanel(
      leafletOutput("map", height = "700px")
    )
  )
)


server <- function(input, output, session) {
  output$map <- renderLeaflet({
    mapview(
      wildfire_sf,
      zcol = input$var,
      layer.name = input$var
    )@map
  })
}

shinyApp(ui, server)
```

## 🧩 Challenge 5

Can you add the functionality to your census shiny app (new panel) that allows users to pick two census variables to plot on an interactive scatter plot using `ggplotly()` and `stat_cor()`

We should remain careful about our use/potential dependence on AI for coding but for cases like these where we have built all the individual parts, AI can be really helpful in putting them together in a sleek design on a shiny app.

```{r}
library(shiny)
library(leaflet)
library(mapview)
library(sf)
library(ggplot2)
library(plotly)
library(ggpubr)  # for stat_cor()

# --- Data setup ---

wildfire_sf <- st_as_sf(
  wildfire_df,
  coords = c("longitude", "latitude"),
  crs = 4326
)

map_variable_choices <- c("daynight", "scan", "track")

# hawaii_data is already sf — no conversion needed
census_variable_choices <- c(
  "total_population",
  "bachelors_count",
  "median_income",
  "median_gross_rent",
  "pct_bachelors"
)

# --- UI ---
ui <- fluidPage(
  titlePanel("Hawaii Census & Wildfire Explorer"),
  tabsetPanel(
    
    # ---- Existing map panel ----
    tabPanel(
      "Wildfire Map",
      sidebarLayout(
        sidebarPanel(
          selectInput("var", "Choose variable to map", choices = map_variable_choices)
        ),
        mainPanel(
          leafletOutput("map", height = "700px")
        )
      )
    ),
    
    # ---- New scatter plot panel ----
    tabPanel(
      "Census Scatter Plot",
      sidebarLayout(
        sidebarPanel(
          selectInput("xvar", "X variable", choices = census_variable_choices, selected = "median_income"),
          selectInput("yvar", "Y variable", choices = census_variable_choices, selected = "pct_bachelors")
        ),
        mainPanel(
          plotlyOutput("scatter", height = "600px")
        )
      )
    )
  )
)

# --- Server ---
server <- function(input, output, session) {
  
  output$map <- renderLeaflet({
    mapview(
      wildfire_sf,
      zcol = input$var,
      layer.name = input$var
    )@map
  })
  
  output$scatter <- renderPlotly({
    p <- ggplot(hawaii_data, aes(x = .data[[input$xvar]], y = .data[[input$yvar]])) +
      geom_point(alpha = 0.6, color = "hotpink") +
      geom_smooth(method = "lm", se = FALSE, color = "purple") +
      stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top") +
      theme_minimal() +
      labs(
        x = input$xvar,
        y = input$yvar,
        title = paste(input$xvar, "vs", input$yvar)
      )
    
    ggplotly(p)
  })
}

shinyApp(ui, server)

#Please note this whole chunk is AI, I wanted to see how it would actually look like since my original codes kept running into errors and didn't work. I ran out of time, sorry. I just changed up colors, height, and sizes. Thank you. 
```
